Solving Collective Spin Hamiltonians with Decoherence

### 2018-09-28 soc_squeezing/cumulant_expansions.tex
removed unnecessary brackets
fixed a few minor errors and clarified a bit of text

The case of the one-axis twisting is simplest to consider, as we can
write the Hamiltonian in the form
\begin{align}
  H_{\t{OAT}} = \f14 \chi \sum_{j,k} \sigma_\z^{(j)} \sigma_\z^{(k)}
  = \f12 \chi \sum_{j<k} \sigma_\z^{(j)} \sigma_\z^{(k)} + \f14 N \chi,
\end{align}
which admits an exact solution previously worked out by Foss-Feig et
al.~in ref.~\cite{foss-feig2013nonequilibrium}.  Adapting exact
expectation values with respect to the initial state $\ket{\X}$
satisfying $S_\x\ket{\X}=S\ket{\X}$, in terms of $S\equiv N/2$ and
$\mu,\nu\in\set{+1,-1}$ we find that
\begin{align}
  \bk{S_+}
  &= \sum_j\bk{\sigma_+^{(j)}}
  = S e^{-\Gamma t} \Phi\p{\chi,t}^{N-1}, \label{eq:S+_OAT} \\
  \bk{S_\mu S_\z}
  &= \f12\sum_j\bk{\sigma_\mu^{(j)}\sigma_\z^{(j)}}
  + \f12\sum_{j\ne k} \bk{\sigma_\mu^{(j)} \sigma_\z^{(k)}} \\
  &= -\f{\mu}{2}\bk{S_\mu} + S \p{S-\f12} e^{-\Gamma t}
  \Psi\p{\mu\chi,t} \Phi\p{\chi,t}^{N-2}, \\
  \bk{S_\mu S_\nu}
  &= \sum_j \bk{\sigma_\mu^{(j)} \sigma_\nu^{(j)}}
  + \sum_{j\ne k} \bk{\sigma_\mu^{(j)} \sigma_\nu^{(k)}} \\
  &= \delta_{\mu,-\nu} \p{S + \mu\bk{S_\z}}
  + S \p{S-\f12} e^{-2\Gamma t}
  \Phi\p{\sp{\mu+\nu}\chi,t}^{N-2}, \label{eq:SS+-_OAT}
\end{align}
where
\begin{align}
  \Phi\p{X,t}
  &\equiv e^{-\lambda t} \sp{\cos\p{t\sqrt{s_X^2-r}}
    + \lambda t~\t{sinc}\p{t\sqrt{s_X^2-r}}},
  \\
  \Psi\p{X,t}
  &\equiv e^{-\lambda t} \p{is_X-\gamma}t~
  \t{sinc}\p{t\sqrt{s_X^2-r}},
\end{align}
for
\begin{align}
  \gamma \equiv \f12 \p{\gamma_+ - \gamma_-},
  &&
  \lambda \equiv \f12 \p{\gamma_+ + \gamma_-},
  &&
  r \equiv \gamma_+ \gamma_-,
  &
  \Gamma \equiv \gamma_\z + \lambda,
  &&
  s_X \equiv X + i\gamma,
\end{align}
defined respectively in terms of spontaneous uncorrelated spin
dephasing, excitation, and decay rates $\gamma_\z$, $\gamma_+$,
$\gamma_-$.  The last ingredients we need to compute spin squeezing at
any time are $\bk{S_\z}$ and $\bk{S_\z^2}$.  As these operators
commute with the OAT Hamiltonian, their evolution is governed entirely
by decoherence (see Appendix \ref{sec:decay_single}), which means
\begin{align}
  \f{d}{dt} S_\z
  &= S\p{\gamma_+-\gamma_-} - \p{\gamma_++\gamma_-} S_\z,
  \\
  \f{d}{dt}\p{S_\z^2}
  &= S\p{\gamma_++\gamma_-} + 2\p{\gamma_+-\gamma_-}\p{S-\f12} S_\z
  - 2 \p{\gamma_++\gamma_-} S_\z^2.
\end{align}
The initial conditions $\bk{S_\z}_{t=0}=0$ and $\bk{S_\z^2}_{t=0}=S/2$
then imply
\begin{align}
  \bk{S_\z}
  = S\p{\f{\gamma_+-\gamma_-}{\gamma_++\gamma_-}}
  \p{1-e^{-\p{\gamma_+ + \gamma_-} t}},
  &&
  \bk{S_\z^2} = S \sp{\f12 + \p{S-\f12} \f{\bk{S_\z}^2}{S^2}}.
  \label{eq:Sz_OAT}
\end{align}
The expectation values in \eqref{eq:S+_OAT}-\eqref{eq:SS+-_OAT} and
\eqref{eq:Sz_OAT} are sufficient to compute the squeezing parameter
$\xi^2$ defined in \eqref{eq:squeezing} at any time throughout
evolution of the initial state $\ket{\X}:S_\x\ket{\X}=S\ket{\X}$ under
$H_{\t{OAT}}$.

@@

The time derivative operator $d/dt$ will generally couple spin
operators $S_+^\ell S_\z^m S_-^n$ to spin operators
$S_+^p S_\z^q S_-^r$ with higher weight, i.e.~with $p+q+r>\ell+m+n$.
In practice, keeping track of this operator growth will eventually
require more computational resources than are available, so we must
truncate it at some point by taking
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_\z^m S_-^n}
  = \sum_{p,q,r} T^{\ell m n}_{pqr} S_+^p S_\z^q S_-^r
  \to \sum_{w\p{p,q,r}<W} T^{\ell m n}_{pqr} S_+^p S_\z^q S_-^r
  \label{eq:time_derivative_truncation}
\end{align}
for some cutoff $W$ and weight function $w$, e.g.~$w\p{p,q,r}=p+q+r$.
The truncation in \eqref{eq:time_derivative_truncation} will close the
system of differential equations defined by
\eqref{eq:time_derivative}, and allow us to solve it using standard
numerical methods.  The initial conditions for this system of
differential equations, i.e.~expectation values of operators with
respect to states that are simple to prepare, are provided in Appendix
\ref{sec:initial_conditions}.

The truncation strategy in \eqref{eq:time_derivative_truncation} has a
few limitations:
\begin{enumerate*}[label=(\roman*)]
\item simulating a system of differential equations for a large number
  of operators can be very time-consuming,
\item the weight function $w$ must be chosen carefully, as the optimal
  weight function is generally system-dependent,
\item simulation results can only be trusted up to the time at which
  the initial value of operators $S_+^p S_\z^r S_-^q$ with weight
  $w\p{p,q,r}\ge W$ have a non-negligible contribution to operators of
  interest.\label{pt:limitation}
\end{enumerate*}
While the first two of these limitations might in principle be avoided
by alternate means of tracking operators, the last limitation
necessarily applies in some form to any method tracking only a subset
of {\it all} operators.  Embracing the unavoidable, we therefore try
to devise an alternate truncation strategy which is built around
\ref{pt:limitation}.

@@

  &= \f1{r!} \sum_{\p{\v k,p-r,\ell}}

@@

We will generally be concerned with ordered products of collective
spin operators of the form $S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n$ for
$\bmu\equiv-\mu\in\set{+1,-1}$.  The most general (anti-)commutator we
need to consider is then
\begin{align}
  \sp{S_\mu^p \p{\mu S_\z}^q S_\bmu^r,
    S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n}_\pm
  = \S^{pqr}_{\ell mn;\mu} \pm \S^{\ell mn}_{pqr;\mu},
\end{align}
where
\begin{align}
  \S^{pqr}_{\ell mn;\mu}
  = S_\mu^p \p{\mu S_\z}^q S_\bmu^r
  S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n
  = \sum_{k=0}^{\min\set{r,\ell}} k! { r \choose k } { \ell \choose k }
  S_\mu^{p+\ell-k} \tilde Z_{qr\ell m;\mu}^{(k)} S_\bmu^{r+n-k},
  \label{eq:general_product}
\end{align}
for
\begin{align}
  \tilde Z_{qr\ell m;\mu}^{(k)}
  &\equiv \p{\sp{\ell-k}+\mu S_\z}^q
  Z_{r\ell;\mu}^{(k)} \p{\sp{r-k}+\mu S_\z}^m \\
  &= \sum_{a=0}^k \zeta_{r\ell}^{ka}
  \sum_{b=0}^q \p{\ell-k}^{q-b} { q \choose b }
  \sum_{c=0}^m \p{r-k}^{m-c} { m \choose c }
  \p{\mu S_\z}^{a+b+c},
\end{align}
defined in terms of $Z_{r\ell;\mu}^{(k)}$ and $\zeta_{r\ell}^{ka}$ as
given in \eqref{eq:Z_mnp} and \eqref{eq:zeta_mnpq}.

>>>>

The case of the one-axis twisting is simplest to consider, as we can
write the Hamiltonian in the form
\begin{align}
  H_{\t{OAT}} = \f14 \chi \sum_{j,k} \sigma_\z^{(j)} \sigma_\z^{(k)}
  = \f12 \chi \sum_{j<k} \sigma_\z^{(j)} \sigma_\z^{(k)} + \f14 N \chi,
\end{align}
which admits an exact solution previously worked out by Foss-Feig et
al.~in ref.~\cite{foss-feig2013nonequilibrium}.  Adapting exact
expectation values with respect to the initial state $\ket{\X}$
satisfying $S_\x\ket{\X}=S\ket{\X}$, in terms of $S\equiv N/2$ and
$\mu,\nu\in\set{+1,-1}$ we find that
\begin{align}
  \bk{S_+}
  &= \sum_j\bk{\sigma_+^{(j)}}
  = S e^{-\Gamma t} \Phi\p{\chi,t}^{N-1}, \label{eq:S+_OAT} \\
  \bk{S_\mu S_\z}
  &= \f12\sum_j\bk{\sigma_\mu^{(j)}\sigma_\z^{(j)}}
  + \f12\sum_{j\ne k} \bk{\sigma_\mu^{(j)} \sigma_\z^{(k)}} \\
  &= -\f{\mu}{2}\bk{S_\mu} + S \p{S-\f12} e^{-\Gamma t}
  \Psi\p{\mu\chi,t} \Phi\p{\chi,t}^{N-2}, \\
  \bk{S_\mu S_\nu}
  &= \sum_j \bk{\sigma_\mu^{(j)} \sigma_\nu^{(j)}}
  + \sum_{j\ne k} \bk{\sigma_\mu^{(j)} \sigma_\nu^{(k)}} \\
  &= \delta_{\mu,-\nu} \p{S + \mu\bk{S_\z}}
  + S \p{S-\f12} e^{-2\Gamma t}
  \Phi\p{\sp{\mu+\nu}\chi,t}^{N-2}, \label{eq:SS+-_OAT}
\end{align}
where
\begin{align}
  \Phi\p{X,t}
  &\equiv e^{-\lambda t} \sp{\cos\p{t\sqrt{s_X^2-r}}
    + \lambda t~\t{sinc}\p{t\sqrt{s_X^2-r}}},
  \\
  \Psi\p{X,t}
  &\equiv e^{-\lambda t} \p{is_X-\gamma}t~
  \t{sinc}\p{t\sqrt{s_X^2-r}},
\end{align}
for
\begin{align}
  \gamma \equiv \f12 \p{\gamma_+ - \gamma_-},
  &&
  \lambda \equiv \f12 \p{\gamma_+ + \gamma_-},
  &&
  r \equiv \gamma_+ \gamma_-,
  &&
  \Gamma \equiv \gamma_\z + \lambda,
  &&
  s_X \equiv X + i\gamma,
\end{align}
defined respectively in terms of spontaneous uncorrelated spin
dephasing, excitation, and decay rates $\gamma_\z$, $\gamma_+$,
$\gamma_-$.  The last ingredients we need to compute spin squeezing at
any time are $\bk{S_\z}$ and $\bk{S_\z^2}$.  As these operators
commute with the OAT Hamiltonian, their evolution is governed entirely
by decoherence (see Appendix \ref{sec:decay_single}), which means
\begin{align}
  \f{d}{dt} S_\z
  &= S\p{\gamma_+-\gamma_-} - \p{\gamma_++\gamma_-} S_\z,
  \\
  \f{d}{dt}\p{S_\z^2}
  &= S\p{\gamma_++\gamma_-} + 2\p{\gamma_+-\gamma_-}\p{S-\f12} S_\z
  - 2 \p{\gamma_++\gamma_-} S_\z^2.
\end{align}
The initial conditions $\bk{S_\z}_{t=0}=0$ and $\bk{S_\z^2}_{t=0}=S/2$
then imply
\begin{align}
  \bk{S_\z}
  = S\p{\f{\gamma_+-\gamma_-}{\gamma_++\gamma_-}}
  \p{1-e^{-\p{\gamma_+ + \gamma_-} t}},
  &&
  \bk{S_\z^2} = S \sp{\f12 + \p{S-\f12} \f{\bk{S_\z}^2}{S^2}}.
  \label{eq:Sz_OAT}
\end{align}
The expectation values in \eqref{eq:S+_OAT}-\eqref{eq:SS+-_OAT} and
\eqref{eq:Sz_OAT} are sufficient to compute the squeezing parameter
$\xi^2$ defined in \eqref{eq:squeezing} at any time throughout
evolution of the initial state $\ket{\X}:S_\x\ket{\X}=S\ket{\X}$ under
$H_{\t{OAT}}$.

@@

The time derivative operator $d/dt$ will generally couple spin
operators $S_+^\ell S_\z^m S_-^n$ to spin operators
$S_+^p S_\z^q S_-^r$ with higher ``weight'', i.e.~with
$p+q+r>\ell+m+n$, which signifies the growth of many-body
correlations.  In practice, keeping track of this operator growth will
eventually require more computational resources than are available, so
we must truncate it at some point by taking
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_\z^m S_-^n}
  = \sum_{p,q,r} T^{\ell m n}_{pqr} S_+^p S_\z^q S_-^r
  \to \sum_{w\p{p,q,r}<W} T^{\ell m n}_{pqr} S_+^p S_\z^q S_-^r
  \label{eq:time_derivative_truncation}
\end{align}
for some cutoff $W$ and weight function $w$, e.g.~$w\p{p,q,r}=p+q+r$.
The truncation in \eqref{eq:time_derivative_truncation} will close the
system of differential equations defined by
\eqref{eq:time_derivative}, and allow us to solve it using standard
numerical methods.  The initial conditions for this system of
differential equations, i.e.~expectation values of operators with
respect to states that are simple to prepare, are provided in Appendix
\ref{sec:initial_conditions}.

The truncation strategy in \eqref{eq:time_derivative_truncation} has a
few limitations:
\begin{enumerate*}[label=(\roman*)]
\item simulating a system of differential equations for a large number
  of operators can be very time-consuming,
\item the weight function $w$ may need to be chosen carefully, as the
  optimal weight function is generally system-dependent,
\item simulation results can only be trusted up to the time at which
  the initial value of operators $S_+^p S_\z^r S_-^q$ with weight
  $w\p{p,q,r}\ge W$ have a non-negligible contribution to operators of
  interest.\label{pt:limitation}
\end{enumerate*}
While the first two of these limitations might in principle be avoided
by alternate means of tracking operators, the last limitation
necessarily applies in some form to any method tracking only a subset
of {\it all} operators.  Embracing the unavoidable, we therefore try
to devise an alternate truncation strategy which is built around
\ref{pt:limitation}.

@@

  &= \f1{\p{r-1}!} \sum_{\p{\v k,p-r,\ell}}

@@

We will generally be concerned with ordered products of collective
spin operators of the form $S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n$ for
$\bmu\equiv-\mu\in\set{+1,-1}$.  The most general (anti-)commutator we
need to consider is then
\begin{align}
  \sp{S_\mu^p \p{\mu S_\z}^q S_\bmu^r,
    S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n}_\pm
  = \S^{pqr}_{\ell mn;\mu} \pm \S^{\ell mn}_{pqr;\mu},
\end{align}
where
\begin{align}
  \S^{pqr}_{\ell mn;\mu}
  = S_\mu^p \p{\mu S_\z}^q S_\bmu^r
  S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n
  = \sum_{k=0}^{\min\set{r,\ell}} k! { r \choose k } { \ell \choose k }
  S_\mu^{p+\ell-k} \tilde Z_{qr\ell m;\mu}^{(k)} S_\bmu^{r+n-k},
  \label{eq:general_product}
\end{align}
for
\begin{align}
  \tilde Z_{qr\ell m;\mu}^{(k)}
  &\equiv \p{\ell-k+\mu S_\z}^q
  Z_{r\ell;\mu}^{(k)} \p{r-k+\mu S_\z}^m \\
  &= \sum_{a=0}^k \zeta_{r\ell}^{ka}
  \sum_{b=0}^q \p{\ell-k}^{q-b} { q \choose b }
  \sum_{c=0}^m \p{r-k}^{m-c} { m \choose c }
  \p{\mu S_\z}^{a+b+c},
\end{align}
defined in terms of $Z_{r\ell;\mu}^{(k)}$ and $\zeta_{r\ell}^{ka}$ as
given in \eqref{eq:Z_mnp} and \eqref{eq:zeta_mnpq}.
